Parsing Postfix log files

### 2008-04-15 doc/logparser.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

\newpage
\section{Introduction}

\label{introduction}

Most mail server administrators will have performed some basic processing
of the logs produced by their mail server at one time or another, whether
it was to debug a problem, explain to a user why their mail is being
rejected, or check whether new anti-spam measures are working.  The more
adventurous will have generated statistics to show how many hits each of
their anti-spam measures has gotten in the last week, and possibly even
generated some graphs to clearly illustrate the point to management or
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing log lines
independently.  One of the barriers to this type of processing is the
unstructured nature of Postfix logs, where each log line was added on an ad
hoc basis as a requirement was discovered or new functionality was
added.\footnote{A history of all changes made to Postfix is distributed
with the source code, available from \url{http://www.postfix.org/}} Further
complication arises because the set of rejection messages is not fixed: new
messages can be added by the administrator with custom checks; every
\RBL{}\footnote{This document is supplied with a glossary, see
appendix~\ref{Glossary}.} returns a different explanatory message; policy
servers may log different messages depending on the characteristics of the
connection; there are many ways in which the log lines may differ between
servers, even within the same organisation --- servers may be configured
differently, or running different version of Postfix.  This paper documents
the difficult process of parsing Postfix logs, presenting \PLP{}, a program
which parses logs and places the resulting data into a database for later
use.  The gathered data can then be used to optimise current anti-spam
measures, provide a baseline to test new anti-spam measures against, or to
produce statistics showing how effective those measures are.  There are
numerous other uses for such data: improving server performance by
identifying troublesome destinations and reconfiguring appropriately;
identifying regular high volume uses (e.g.\ customer newsletters) and
restricting those uses to off-peak times; detecting virus outbreaks which
propagate via email; as a base for billing customers on a shared server.
Preserving the raw data enables users to develop a multitude of uses far
beyond those conceived of by the author.

\vspace{1em}\noindent\textbf{Layout of the paper:}

Section~\ref{background} provides background information useful in
understanding the paper, parser and algorithm.  It introduces the idea of
using a database schema as an \API{}, providing an interface to the data
gathered that is language-neutral.  The unusual separation of rules,
actions and algorithm is discussed, giving the reasons that approach was
taken when designing the parser.

This algorithm requires a database for storing both the rules used when
parsing and the results gleaned from parsing.  The database schema used is
described in section~\refwithpage{database schema}, explaining in detail
the tables used for storing the data gleaned from the logs and the table
that stores the rules.

Section~\ref{rules} discusses the parsing rules in detail, explaining the
purpose and usage of each field in a rule, referring to an example rule and
sample data it matches successfully against.  The pros and cons of
overlapping rules are discussed, including techniques for detecting
unintentional overlaps.  Rule efficiency concerns are discussed, in
particular the optimisations used by the algorithm, with reference to the
graphs in appendix~\refwithpage{graphs}.  The final section is a
description of using the tools provided with the parser to generate new
rules (specifically the \regex{} in each rule) from unparsed log lines.

Section~\ref{parsing-algorithm} contains the core of the paper, describing
a naive parsing algorithm and the complications initially encountered which
shaped the full algorithm.  A flow chart and a discussion of the emergent
behaviour exhibited by the algorithm accompanies a comprehensive
explanation of the different stages of the initial algorithm.  The
framework which actions and rules fit into is documented, then the actions
taken during execution of the algorithm are described, followed by the
process of adding a new action.  The section concludes with an in-depth
description of the further complications discovered and their solutions
which complete the parser.

Section~\ref{parsing coverage} analyses the coverage the parser achieves
over a set of \numberOFlogFILES{} log files taken from a mail server
handling mail for over 700 users, averaging 8500 mails per day
(graph~\refwithpage{Mails received per day}).  Coverage is described both
in terms of the fraction of log lines parsed and the fraction of mails and
connections successfully reconstructed by the parsing algorithm, including
dealing with false negatives and a discussion of the difficulties in
identifying false positives.  A random sampling of log lines was parsed,
and the correctness of the results manually verified, as part of
determining the coverage of the parser.

Section~\ref{limitations-improvements} lists the limitations of the
algorithm, then suggests some ways of dealing with them, with the goal of
improving parsing and reproduction of the journey a mail takes through
Postfix.

Section~\ref{conclusion} contains the paper's conclusion, describing the
results of the research, design and implementation of the parser.

Appendix~\ref{other-parsers} discusses other parsers and why they were
deemed unsuitable for the task, including why they could not be improved or
expanded upon.

The bibliography contains references to the resources used in developing
the algorithm, writing the program, and preparing this paper.  Also listed
are some additional resources expected to be helpful in understanding
\SMTP{}, Postfix, anti-spam techniques, or the paper.

Appendix~\ref{graphs} contains graphs illustrating the topics previously
discussed in \sectionref{rule efficiency} (rule efficiency), and explains
the irregularities observed in these graphs.

Appendix~\ref{Glossary} provides a glossary of terms used in the paper.

Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.

\section{Background}

\label{background}

\subsection{Introduction}

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
introduction to Postfix, concentrating on the topics most relevant to this
paper, namely Postfix anti-spam restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this paper.  Other projects which attempt to
parse Postfix logs are summarised (full details are available in
appendix~\refwithpage{other-parsers}), finishing with a review of
previously published research in this area.

\subsection{Motivation}

\label{motivation}

This paper and the program it describes are part of a larger project to
optimise a server's Postfix restrictions, generate statistics and graphs,
and provide a platform on which new restrictions can be trialled and
evaluated to see if they are beneficial in the fight against spam.  The
program parses Postfix logs and populates a database with the data gleaned
from those logs, providing a consistent and simple view of the logs which
future tools can utilise.  The gathered data can then be used to optimise
current anti-spam measures, provide a baseline to test new anti-spam
measures against, or to produce statistics showing how effective those
measures are.

A short example of the optimisation possible using data from the database
is determining which Postfix restrictions reject the highest number of
mails:

\begin{verbatim}
SELECT name, description, restriction_name, hits_total
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY hits_total DESC;
\end{verbatim}

If the database supports sub-selects percentages can be obtained:
\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
the database containing the extracted data does not support this syntax,
then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
results will be the same, just slightly less visually pleasing.}

\begin{verbatim}
SELECT name, description, restriction_name, hits_total,
        (hits_total * 100.0 /
            (SELECT SUM(hits_total)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY hits_total DESC;
\end{verbatim}

Another example is determining which restrictions are not effective: this
shows which restrictions had fewer than 100 hits on the last log file
parsed.

\begin{verbatim}
SELECT name, description, restriction_name, hits,
        (hits * 100.0 /
            (SELECT SUM(hits)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
        AND hits < 100
    ORDER BY hits ASC;
\end{verbatim}

These database queries yield summary statistics about the efficiency of
spam avoidance techniques that is far less feasible to assess directly from
log files without prior pre-processing into a database in the fashion
proposed, implemented and tested herein.

\subsection{Database as Application Programming Interface}

\label{database as API}

The database populated by this program provides a simple interface to
Postfix logs.  Although the interface is a database schema, it is in effect
quite similar to any other \API{} provided by shared code: it insulates
both user and provider of the \API{} from changes in the implementation of
the \API{}\@.  The algorithm implemented by the parser can be improved;
support can be added for earlier or later releases of Postfix; bugs can be
fixed or limitations removed from the parser; these changes will not cause
the user to be negatively impacted.  Statistics and/or graphs can be
generated from the database; new restrictions can be tested and the results
inspected; trends in the fight against spam can emerge from historical data
saved in the database; the parser remains the same as the usage adapts.
Using a database simplifies writing programs which need to interact with
the data in several ways:

    \item The majority of programming languages have existing code
        available providing access to databases, so the majority of
        languages will be able to access the results obtained by the
        parser.  If this parser was written to be utilised as shared code
        there would be little or no difficulty in using it when writing in
        the programing language it is implemented in, but every other
        programing language wishing to utilise it would require an
        interface layer to be written.  This requirement would drastically
        reduce the viability of any project wishing to build upon the work
        of this project.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of variable complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

    \item Databases are easily extensible, e.g.:

        \begin{itemize}

            \item Other tables can be added to the database, e.g.\ to cache
                historical, summary or computed data.

            \item New columns can be added to the tables used by the
                program, with sufficient DEFAULT clauses or a clever
                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
                for explanations of these terms,
                e.g.~\cite{sql-for-web-nerds}}

            \item A VIEW gives a custom arrangement of data with very
                little effort.

            \item If the database supports it, access can be granted on a
                fine-grained basis so that the finance department can
                produce invoices, the helpdesk can run limited queries as
                part of dealing with support calls, and the administrators
                can have full access to the data.

            \item Triggers can be written to perform actions when certain
                events occur.  In pseudo-\SQL{}\@:

\begin{verbatim}
CREATE TRIGGER ON INSERT INTO results
    WHERE sender = 'boss@example.com'
        AND postfix_action = 'REJECTED'
    SEND PANIC EMAIL TO 'postmaster@example.com';
\end{verbatim}

        \end{itemize}

    \item \SQL{} is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn (a good
        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
        every vendor implements a different dialect of \SQL{}, the basics
        are the same everywhere (analogous to the overall similarities and
        minor differences between Irish English, British English, American
        English and Australian English).  Depending on the database in use
        there may be tools available which reduce or remove the requirement
        to know \SQL{}; for SQLite (the default database used by the
        implementation) there are several available~\cite{sqlite-guis}.

Storing the results in a database will also increase the efficiency of
using those results, as the logs need only be parsed once; indeed the
results may be used by someone with no access to the original logs.

\subsection{Simple Mail Transfer Protocol}
\label{SMTP background}

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and
viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
adhered to the principle of \textit{Be liberal in what you accept, and
conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
unfortunately that principle was written in a friendlier time.  Given the
deluge of spam that mail servers are subjected to daily, a more appropriate
maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
the strongest restrictions you can; relax the restrictions and adherence
only when legitimate mail is impeded.\/}  It's not as friendly, nor as
catchy, but it more accurately reflects the current situation.} A typical
\SMTP{} conversation resembles the following (the lines starting with a
three digit number are sent by the server, all other lines are sent by the
client):

\begin{verbatim}
220 smtp.example.com ESMTP
HELO client.example.com
250 smtp.example.com
MAIL FROM: <alice@example.com>
250 2.1.0 Ok
RCPT TO: <bob@example.com>
250 2.1.5 Ok
DATA
354 End data with <CR><LF>.<CR><LF>
Message headers and body sent here.
.
250 2.0.0 Ok: queued as D7AFA38BA
QUIT
221 2.0.0 Bye
\end{verbatim}

An example deviation from the protocol:

\begin{verbatim}
220 smtp.example.com ESMTP
HELO client.example.com
250 smtp.example.com
MAIL FROM: Alice N. Other alice@example.com
501 5.1.7 Bad sender address syntax
RCPT TO: Bob in Sales/Marketing bob@example.com
503 5.5.1 Error: need MAIL command
DATA
503 5.5.1 Error: need RCPT command
Message headers and body sent here.
.
502 5.5.2 Error: command not recognized
QUIT
221 2.0.0 Bye
\end{verbatim}

This client is so poorly written that not only does it present the sender
and recipient addresses improperly, it ignores the error messages returned
by the server and carries on regardless.  There are many spam and virus
sending programs which are this deficient --- unfortunately others
(particularly newer programs) were written by competent programmers, or
utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
would have done its best to deal with deficient clients, with the intention
of accepting as much mail destined for its users as
possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
command, or accepting sender or recipient addresses which were not enclosed
in \texttt{<>}.

A detailed description of \SMTP{} is beyond the scope of this document:
introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.

\subsection{Postfix}

\label{postfix background}

Postfix is a \MTA{} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive, extensible, optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of severity and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained in \sectionref{policy
servers}.} as the ultimate in flexibility.  Administrators can also supply
their own rejection messages to make it clear to senders why exactly their
mail was rejected.  Unfortunately this flexibility has a cost: complexity
in the logs generated.  While it is easy to use standard Unix text
processing utilities to determine the fate of an individual email,
following the journey an email takes through Postfix can be quite
difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
the journey is simple, but the other 10\% of the time requires 90\% of the
code.\footnote{These numbers don't have a solid scientific basis, they're
based on gut feeling from writing and debugging the software.}

Postfix's design follows the Unix philosophy of \textit{Write programs that
do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
principle was also followed when designing the parser (\sectionref{parser
design}), notably having separate actions rather than one huge action.} and
is separated into various component programs to perform the tasks required
of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
full details can be found at~\cite{postfix-overview}.  Each log line
contains the name of the Postfix program which produced it, and this
information is used when determining which rules should be used to parse
each log line (see \sectionref{rule characteristics} for details).  This
design has positive security implications also: those components which
interact with other hosts are not privileged,\footnote{Privilege in this
case means the power to perform actions that are limited to the
administrator, and not available to ordinary users.} so bugs in those
components will not give an attacker extra privileges; those components
which are privileged do not interact with other hosts, making it much more
difficult for an attacker to exploit any bugs which may exist in those
components.

\subsubsection{Mixing and matching Postfix restrictions}

Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
smtpd_per_user_control, policy-servers}, the following is a brief overview
only.

Postfix uses one restriction list (containing zero or more restrictions)
for each stage of the \SMTP{} conversation: client connection, HELO
command, MAIL FROM command, RCPT TO commands, DATA command, and end of
data.  The appropriate restriction list is evaluated for each stage
(evaluation will be explained shortly), though by default the restriction
lists for client connection, HELO and MAIL FROM commands will not be
evaluated until the first RCPT TO command is received, because some clients
don't deal properly with rejections before RCPT TO\@; a benefit of this
delay is that Postfix has more information available when logging the
rejection.  Postfix uses simple lookup tables as the deciding factor when
evaluating some restrictions, e.g.\newline
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\newline The line above can be broken down as follows:

\begin{description}

    \item [check\_client\_access] The name of the restriction to evaluate.

    \item [cidr] The type of the lookup table.

    \item [/etc/postfix/client\_access] The file containing the lookup
        table.

\end{description}

The restriction \texttt{check\_client\_access} checks whether the \IP{}
address of the connected client is found in the file and returned the
associated action if found; the method of searching the file is dependant
on the type of the file (\texttt{cidr} in the example) --- see
\cite{postfix-lookup-tables} for more details.  Other restrictions
determine their result by consulting external sources, e.g.\
\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
address of the client is present in the \RBL{} \texttt{rbl.example.com},
rejecting the command if the client is listed.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue processing the remainder of the restriction list, allowing more
specific cases to be used as exceptions to more general cases.  When the
result is the name of another restriction Postfix will evaluate the new
restriction, allowing restrictions to be chosen based on the client \IP{}
address, HELO hostname, sender address, recipient address, etc.  The
administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex sequences of lookups,
restrictions and exceptions.  Postfix tries to protect the administrator in
as far as is reasonable, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
all users --- the administrator must remember that \textit{With great
power comes great responsibility}.

This description is necessarily brief, for further details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.

\subsubsection{Policy servers}

\label{policy servers}

A policy server~\cite{policy-servers} is an external program that accepts
state information from Postfix for each \SMTP{} command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  For example some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent problems from
spam or (worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

\begin{itemize}

    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
        \SPF{}\label{spf introduction} records specify which mail servers
        are allowed to send mail claiming to be from a particular domain.
        The intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and
        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects mail when the triple of (sender address, recipient address,
        remote \IP{} address) is unknown; on second and subsequent delivery
        attempts from that triple the mail will be accepted.  The
        assumption is that maintaining a list of failed addresses and
        retrying after a temporary failure is uneconomical for a spam
        sender, but that a legitimate mail server must retry.  Sadly spam
        senders are using increasingly complex and well written programs to
        distribute spam, frequently using an \ISP{} provided \SMTP{} server
        from a compromised machine on the \ISP{}'s network.  Greylisting
        will slowly become less useful, but it does block a large
        percentage of spam mail at the moment; the most effective
        restrictions over the \numberOFlogFILES{} log files used in testing
        the parser are shown in table~\refwithpage{Summary of rejections}.
        Greylisting is obviously worth using, at least at the moment,
        particularly when you factor in Greylisting's position as the final
        restriction which a mail must overcome:\footnote{Greylisting is the
        final restriction a mail must overcome in the configuration used on
        the mail server the logs were obtained from; an administrator is
        free to use Greylisting at whichever position in the restriction
        list they feel is most appropriate for their mail system.}
        Greylisting only takes effect for mails which have passed every
        other restriction.

        \begin{table}[ht]
            \caption{Summary of rejections}\label{Summary of rejections}
            \input{build/restriction-table-include.tex}
        \end{table}

    \item Using a scoring system such as
        Policyd-weight~\cite{policyd-weight} where tests accumulate points
        against the sending system --- if the eventual score is too high
        the mail is rejected.  Postfix's restrictions are binary, and the
        administrator must manually whitelist clients if they are to bypass
        a restriction; using a threshold which requires hitting several
        restrictions frees the administrator from whitelisting clients
        which fall foul of one restriction only.

    \item Rate limiting on a per-sender, per-client or per-recipient basis
        as performed by Policyd~\cite{policyd}.

\end{itemize}

Example attributes taken from~\cite{policy-servers}:

\begin{tabular}[]{ll}

    request                 & smtpd\_access\_policy     \\
    protocol\_state         & RCPT                      \\
    protocol\_name          & SMTP                      \\
    helo\_name              & some.domain.tld           \\
    queue\_id               & 8045F2AB23                \\
    sender                  & foo@bar.tld               \\
    recipient               & bar@foo.tld               \\
    recipient\_count        & 0                         \\
    client\_address         & 1.2.3.4                   \\
    client\_name            & another.domain.tld        \\
    reverse\_client\_name   & another.domain.tld        \\
    instance                & 123.456.7                 \\

\end{tabular}

\subsection{Assumptions}

The algorithm described and the program implementing it make a small number
of (hopefully safe and reasonable) assumptions:

\begin{itemize}

    \item The logs are whole and complete: nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging system unable to cope with the volume of
        logs).  On a well run system it is extremely unlikely that any of
        these problems will arise, though it is of course possible,
        particularly when suffering a deluge of spam or a mail loop.

    \item Postfix logs sufficient information to make it possible to
        accurately reconstruct the actions it has taken.  There are a
        number of heuristics used when parsing; see
        \sectionref{identifying-bounce-notifications},
        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
        logging after cleanup} for details.

    \item The Postfix queue has not been tampered with, causing unexplained
        appearance or disappearance of mail.  This may happen if the
        administrator deletes mail from the queue without using 
        \daemon{postsuper}, or if there is filesystem corruption.

\end{itemize}

In some ways this task is similar to reverse engineering or replicating a
black box program based solely on its inputs and outputs.  Although the
source code is available,\footnote{Reading and understanding the source
code would require a significant investment of time; the source code for
Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
box while developing the parser:

\begin{itemize}

    \item The parser is developed using real world logs rather than the
        idealised logs someone would naturally envisage reading the source
        code.

    \item The source code cannot accurately communicate the variety of
        orderings in which log lines are written to the log file, as
        process scheduling independently interferes with logging and other
        processing.

    \item The parser acts as a second source of information, with the
        information gathered from empirical evidence.  An interesting
        project would be to compare the empirical knowledge inherent in the
        parsing algorithm with the documentation and source code of
        Postfix.

\end{itemize}

\subsection{Parser design}

\label{parser design}

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that logs produced by Postfix are not fixed; they vary widely
from host to host, depending on the set of restrictions chosen by the
administrator.  With this in mind, one of the parser's design aims was to
make adding new rules as easy as possible, to enable administrators to
properly parse their log lines.  To enable this the parser is divided into
three parts:

\begin{description}

    \item [rules] Rules match individual log lines and determine which
        actions will be executed.  Rules provide an easily extensible
        method of associating log lines and actions.  Rules are described
        in detail in \sectionref{rules}.

    \item [actions] Actions invoked to deal with a log line once it has been
        identified by the rules: modifying data structures, handing
        complications, and saving data to the database.  The actions
        perform the work of reconstructing the journey a mail takes through
        Postfix.  Full details of actions can be found in
        \sectionref{parsing-algorithm}, particularly
        \sectionref{actions-in-detail} and \sectionref{adding new actions}.

    \item [framework] The framework is responsible for loading rules, managing data
        structures so actions can retrieve them, finding the rule which
        matches each log line, invoking the correct action, etc.  The
        framework provides the structure which actions and rules plug into.
        The framework is not described in detail in this document because
        it is not particularly interesting, but there is some coverage in
        \sectionref{framework}.

\end{description}

It may help to think of the rules and actions as components which plug into
the framework.

\label{why separate rules, actions and framework?}

Decoupling the parsing rules from the associated actions allows new rules
to be written and tested without requiring modifications to the parser
source code (significantly lowering the barrier to entry for new or casual
users who need to parse new log lines), and greatly simplifies both actions
and rules.  Decoupling also creates a clear separation of functionality:
rules handle low level details of identifying log lines and extracting data
from a log line, whereas the actions handle the higher level details of
following the path a mail takes through Postfix, assembling the required
data before storing it, dealing with complications arising, etc.

Decoupling the actions from the framework simplifies both framework and
actions: the framework provides services to the actions, and doesn't need
to deal with the complications which arise, or the task of reconstructing a
mail's journey through Postfix; actions benefit from having services
provided by the framework, freeing them to concentrate on the task of
accurately reconstructing each mail's journey through Postfix and dealing
with the complications described in \sectionref{additional complications}.

Separating the rules from the actions and framework makes it possible to
parse new log lines without modifying the core parsing algorithm.  Although
this may seem like a trivial point, is it substantially more difficult to
understand a program's entire parsing algorithm, identify the correct
location to change and make the appropriate changes, versus adding a new
rule with the action to invoke, a \regex{} to match the log lines, and the
specification of the data to extract.  Bear in mind that the changes must
be made without adversely affecting existing parsing, particularly as there
may be edge cases which are not immediately obvious.\footnote{See
\sectionref{yet-more-aborted-delivery-attempts} for a complication which
occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
changes to the parser's code also complicates upgrades, as the changes
must be preserved during the upgrade, and may clash with changes made by
the developer.\footnote{See \sectionref{adding new actions} for how to add
new actions}  \parsername{} allows the user to add new rules to the
database without changing the parsing algorithm, unless the new log lines
to be parsed require functionality not already provided by the algorithm.
If the new log lines do require new functionality, new actions can be added
to the parser without modifying existing actions or other parts of the
algorithm; only in the rare case that the new actions require support from
other sections of the code will more extensive changes be required.

There is some similarity between the parser's design and William Wood's
\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
creating grammars to parse or generate sentences.  The resemblance between
\ATN{} and the parser is accidental, but it is interesting how two
apparently different approaches share an underlying separation of concerns;
this appears to be a natural division of responsibility and functionality.

% Do Not Reformat!

\begin{tabular}[]{lll}
    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
    Networks            & Algorithm         & Determines the sequence 
                                              of transitions            \\
                        &                   & or actions which 
                                              constitutes a valid       \\
                        &                   & input.                    \\
    Transitions         & Actions           & Save data and impose
                                              conditions the            \\
                        &                   & input must meet to be
                                              considered valid.         \\
    Abbreviations       & Rules             & Responsible for 
                                              classifying input.        \\
\end{tabular}

\subsection{Conventions used in the paper}

The words \textit{connection\/} and \textit{mail\/} are often used
interchangeably in this paper; in general the word used was chosen based on
the context it appears in.

\subsection{Other Postfix log parsers}

Ten other parsers have been reviewed in
appendix~\refwithpage{other-parsers} as part of the background research for
this project.  None of the reviewed parsers perform the type of advanced
parsing and log correlation described here; all are intended to perform a
specific parsing and reporting task, rather than be a generic parser,
extracting data and leaving generation of reports from the data to other
programs.  Some parsers save the data extracted to a data store but the
majority discard all data once they have finished running, making
historical analysis impossible.  The other parsers reviewed all produce a
report of greater or lesser complexity and detail, whereas the program
described here doesn't attempt to produce a report at all; that
responsibility is deferred to a separate program, to be developed later.
The parsing algorithm and program described here are designed to enable
much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

\subsection{Previous research in this area}

\label{prior art}

There only appears to be one prior paper published about parsing Postfix
log files: \textit{Log Mail Analyzer: Architecture and Practical
Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
different from this parser: it attempts to present correlated log data in a
form suitable for a systems administrator to search using the myriad of
standard Unix text processing utilities already available, producing both a
\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
support both \CSV{} and Berkeley DB appears to have been a limiting factor:
\CSV{} is a very simple format where a record is stored in a single line,
with fields separated by a comma or other punctuation symbol.  Problems
with \CSV{} files include the difficulty in escaping separators, providing
multiple values for a field (e.g.\ multiple recipients), adding new fields,
and lack of a bundled schema describing the fields.

Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
\SQL{} databases which store arbitrary tuples; in the main table the key is
an integer referred to by secondary tables, and the value used is a \CSV{}
line containing all of the data for that row.  The secondary by-sender,
by-recipient, by-date and by-\IP{} tables use the
sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
referring to the main table is the value.  In effect this re-implements
\SQL{} foreign keys, but without the functionality offered by even the most
basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
requires custom code to search on some combination of the above, though the
authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
of the three output formats

The data stored is limited to time and date, hostname and \IP{} address of
client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
codes or remote servers\footnote{A single mail may be sent to multiple
remote servers if it was addressed to recipients in different domains, or
Postfix needs to try multiple servers for one or more recipients.} is not
explained in the paper; experimental observation shows that multiple
records are added when there are multiple recipients (sadly the records are
not associated in any way), and presumably the same approach is taken when
there are multiple destination servers.

The schema used with the MySQL database is undocumented, but at least it's
possible to discover the schema with an \SQL{} database, unlike with
Berkeley DB\@; all \SQL{} databases embed the database schema into the
database and provide commands for displaying it.  Berkeley DB does not
embed a schema, as there is neither requirement nor benefit; it only
provides \textbf{(key, value)} pairs, so any additional structuring of the
data is imposed by the application, thus it behoves the application to
document this structure.

Parsing in \LMA{} requires major changes to the code to parse new log lines
or extract additional data.  It does not appear to deal with any of the
complications discussed in \sectionref{complications} (initial
complications) and \sectionref{additional complications} (additional
complications).  It does not differentiate between different types of
rejections, so it is not suitable for the purposes of this project; the
data about which restriction caused the rejection is discarded, whereas an
important goal of this project is to retain that data to aid optimisation
and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
files generated on Solaris hosts because the fields added automatically
differ from those added on Linux hosts; logs from Solaris hosts (and
possibly other operating systems) thus require preprocessing before parsing
by \LMA{}.

\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
DAILY-REJECT initially failed with an error message from the Perl
interpreter; after making some corrections to the code it worked, and
produced the \CSV{} lines for the specified day where the \SMTP{} code
signifies a rejection.  All of these reports are trivially simple to
produce from the \CSV{} file using the standard Unix tool \texttt{awk},
e.g.\ the most complicated, DAILY-REJECT, is merely:

\begin{verbatim}
awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
    { print $0; print " "; }' lma_output.txt
\end{verbatim}

Notes about the command above:

    \item It outputs an unnecessary line containing only a single space
        after each matching line, for compatibility with the output of
        DAILY-REJECT\@.

    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
        line properly.  \LMA{} considers this code to be a rejection, so
        the command above must, too.

    \item In a tiny proportion of records the \SMTP{} code is
        \textit{deferred\/} (again, not a valid code, and most likely due
        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
        so the equivalent \texttt{awk} command must do so, too.

There are a number of differences in the output from DAILY-REJECT and the
\texttt{awk} command:

\begin{itemize}

    \item \LMA{} does not escape or replace the separators used in the
        \CSV{} records when they are present in sender or recipient
        addresses, leading to some mangled records; these records are not
        present in the output from \texttt{awk}.

    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
        produces only one output line, whereas the \texttt{awk} command
        produces all matching lines.

\end{itemize}

In summary, \LMA{} appears to be a proof of concept, written to demonstrate
the point of their paper (that having this information in an accessible
fashion is useful to systems administrators), rather than a program
designed to be extensible and useful in a production environment.

>>>>

\newpage
\input{logparser-introduction.tex}
\input{logparser-background.tex}
\input{logparser-database-schema.tex}
\input{logparser-parsing-rules.tex}
\input{logparser-parsing-algorithm.tex}
\input{logparser-coverage.tex}
\input{logparser-limitations.tex}
\input{logparser-conclusion.tex}
\input{logparser-other-parsers-review.tex}
